Gaussian Process Regression for astronomical time-series

### 2023-04-24 src/tex/ms.tex
[showyourwork] overleaf sync
Minimal corrections. Feel free to cherry pick!

% macros
\newcommand{\apjl}{Astrophysical Journal Letters}
\newcommand{\aj}{Astronomical Journal}
\newcommand{\apj}{Astrophysical Journal}
\newcommand{\apjs}{Astrophysical Journal Supplement}
\newcommand{\pasp}{Publications of the Astronomical Society of the Pacific}
\newcommand{\jgr}{Journal of Geophysical Research}
\newcommand{\aap}{Astronomy and Astrophysics}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\actaa}{Acta Astronomica}
\newcommand{\nat}{Nature}
\newcommand{\prl}{Physical Review Letters}
\newcommand{\prd}{Physical Review D}

@@

We then used the same framework to model the simulated dataset, first fitting only for an excess variance or ``jitter'' term (a method commonly used to account for model mis-specification) and then accounting for the correlated nature of the explicitly using a GP.
For the transit model, we fit for the planet-to-star radius ratio $R_{\rm p}/R_\star$, the mid-transit time $T_0$, the out-of-transit flux $f_0$, and the limb darkening parameters $u_1,\,u_2$.
We then marginalise over either the variance of the jitter term, or the amplitude $\amplitude$ and timescale $\lengthscale$ of the GP noise model.
The posterior constraints on the time of transit and the planet-to-star ratio are shown in the right panel of \autoref{fig:transit}, with the true values of these parameters over-plotted as a black line.
When the correlated noise is absorbed into a jitter term (plotted in orange in \autoref{fig:transit}), the inferred parameters are significantly inconsistent with the truth.
Taking the correlated noise into account (blue in \autoref{fig:transit}) increases the uncertainties on the inferred parameters, but also shifts the results to recover the true parameters.

@@

The kernel function $k$ encodes our beliefs about the stochastic, or random, element of the model, in just the same way as the mean function $m$ encodes our beliefs about the deterministic component of the model. For example, in many circumstances, we would expect that two observations taken close together in time should be more strongly correlated than observations taken further apart, and we would use a decreasing function of the time interval $\dt=|t_i-t_j|$ to represent that. One of the most widely used covariance functions is the squared exponential
\begin{equation}
\label{eq:kse}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,\lengthscale^2}\right).
\end{equation}
This gives rise to smooth (infinitely differentiable) random functions with variance $\amplitude^2$ and characteristic length-scale $\lambda$. We present other commonly used kernel functions and discuss how to select one in \autoref{sec:kernels}.

@@

To address this, \citet{2009ApJ...704...51C} proposed a wavelet-based method to model correlated noise in transit light curves, which can be seen as a special case of GPR with a covariance function belonging to the exponential family, but formulated in terms of wavelets. Rather than modelling the covariance in the time domain, the Power Spectral Density (PSD) of the noise is assumed to be of the form $P(f) \propto f^{-\gamma}$.
Noting that the wavelet transform of such a process gives rise to a nearly diagonal covariance matrix, \citet{2009ApJ...704...51C}  derived an expression for the likelihood that can be computed in $\mathcal{O}(N)$ operations. 
This method has been used widely since its publication to model transit observations from space missions which stare continuously at a given field, such as \textit{Kepler} or TESS.
\begin{armarginnote}
  \entry{Colours of noise}{White, pink and red noise correspond to power spectral densities with power-law indices $\gamma=0$, $1$ and $2$, respectively.}
\end{armarginnote}
However, the use of a wavelet transform requires regular time sampling, which precludes its application to datasets with irregular sampling or significant data gaps, for example from ground-based, space telescopes in low-Earth orbit such as Hubble.

@@

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/RV_framework.pdf}
  \caption{Schematic of the multidimensional GP framework of \citet{2015MNRAS.452.2269R} for modelling activity signals in RV data. The observed RVs (shown as orange points with error bars in the bottom right panel) are modelled alongside observations of activity indicators extracted from the spectra or line profiles (orange points with error bars in the bottom left and bottom middle panels). In all these time-series, the activity signal (shown by the black line in each panel) is modelled as a linear combination of an unobserved (latent) GP, shown in the top-left panel, and its first time-derivative, shown in the top middle panel. In addition, the Keplerian signal(s) of putative planet(s) (the sum of which is shown by the blue lines in the top right and bottom right panels) contribute to the RVs but not to the activity indicators. This hierarchical model is implemented in a Bayesian framework, sampling over all the hyperparameters listed in the blue boxes (subject to appropriate priors) to derive posteriors over the hyperparameters of interest and/or estimate the evidence for models with different numbers of planets. Models with different numbers of planets (or without planets) are then compared using the methods discussed in Section~\ref{sec:model_selection}.}
  \label{fig:RV_framework}
\end{figure}

@@

On the other hand, RV extraction pipelines routinely provide additional indicators measuring the characteristic width and asymmetry of the spectral lines, which are also affected by active regions, as well as spectral activity indicators which trace chromospheric emission in the cores of strong lines. \citet{2015MNRAS.452.2269R} developed a GP framework, schematically illustrated in \autoref{fig:RV_framework}, to model these activity indicators alongside the RVs, and to disentangle them from planetary signals. Each observed time-series is modelled as a linear combination of some latent variable $G(t)$, its time derivative $G'(t)$, white noise, and (in the case of the RVs only) one or more Keplerian signals. $G(t)$ does not have a direct physical interpretation; it loosely corresponds to $F^2$ in the $FF'$ framework of \citet{2012MNRAS.419.3147A}, so that $G'(t) \propto F(t) F'(t)$, but it is modelled as a (typically quasi-periodic) GP, which makes it possible to write down expressions for the covariance between any pair of observations from the any of the time-series included in the analysis. This results in a global covariance matrix of dimensions $(NM,NM)$, where $N$ is the number of observations and $M$ the number of activity indicators included in the analysis (including the RVs themselves). This framework has proved particularly successful for RV confirmation of transiting planets around young stars \citep{2019MNRAS.490..698B,2022MNRAS.512.3060Z}, enabling the detection of planetary signals almost 50 times smaller than the activity signals.  Open-source implementations of this framework have been published \citep{2022MNRAS.509..866B,2022A&A...659A.182D}, and extensions thereof are likely to play a significant role in the analysis of next-generation RV surveys targeting Earth analogues, which need to achieve even better contrast between activity and planetary signals. \citet{2022arXiv220506627C} proposed a GP regression network model for stellar activity signals in RV, showing that it can be used to model RVs of the Sun-as-a-star simultaneously with other activity indicators.

@@

Given the current state of the field, our current recommendation for assessing and mitigating model misspecification is to use the methods discussed in \autoref{sec:model_selection} to determine the suitability of a GP for a specific application.
In the case where there are significant outliers in the dataset, the most mature and commonly used approach is to perform iteratively sigma clip the data as a pre-processing step.
As an alternative, \citet{2015ApJ...812..128C} propose a method for handling non-stationary and structured model misspecification in the context of stellar spectral modelling, by designing a basis of ``local'' kernel functions tuned to capture line shape residuals.
Such domain-specific methods can be developed and validated using the methods discussed in \autoref{sec:model_selection}.

@@

It is reasonably straightforward to implement a simple GP model in code, and within astrophysics it has been common for authors to implement custom GPs for their analysis.
However, things get significantly more complicated when implementing the scalable or approximate methods discussed in the previous section.
Similarly, it can be tedious to experiment with different kernel functions and inference methods without building a non-trivial modelling infrastructure, something that has typically been ad-hoc in astrophysics research.

>>>>

% macros
\newcommand{\apjl}{Astrophysical Journal Letters}
\newcommand{\aj}{Astronomical Journal}
\newcommand{\apj}{Astrophysical Journal}
\newcommand{\apjs}{Astrophysical Journal Supplement}
\newcommand{\pasp}{Publications of the Astronomical Society of the Pacific}
\newcommand{\jgr}{Journal of Geophysical Research}
\newcommand{\aap}{Astronomy and Astrophysics}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\actaa}{Acta Astronomica}
\newcommand{\nat}{Nature}
\newcommand{\prl}{Physical Review Letters}
\newcommand{\prd}{Physical Review D}
\newcommand{\ssr}{Space Science Reviews}

@@

We then used the same framework to model the simulated dataset, first fitting only for an excess variance or ``jitter'' term (a method commonly used to account for model mis-specification) and then accounting for the correlated nature of the noise explicitly using a GP.
For the transit model, we fit for the planet-to-star radius ratio $R_{\rm p}/R_\star$, the mid-transit time $T_0$, the out-of-transit flux $f_0$, and the limb darkening parameters $u_1,\,u_2$.
We then marginalise over either the variance of the jitter term, or the amplitude $\amplitude$ and timescale $\lengthscale$ of the GP noise model.
The posterior constraints on the time of transit and the planet-to-star ratio are shown in the right panel of \autoref{fig:transit}, with the true values of these parameters over-plotted as a black line.
When the correlated noise is absorbed into a jitter term (plotted in orange in \autoref{fig:transit}), the inferred parameters are significantly inconsistent with the truth.
Taking the correlated noise into account (blue in \autoref{fig:transit}) increases the uncertainties on the inferred parameters, but also shifts the results to recover the true parameters.

@@

The kernel function $k$ encodes our beliefs about the stochastic, or random, element of the model, in just the same way as the mean function $m$ encodes our beliefs about the deterministic component of the model. For example, in many circumstances, we would expect that two observations taken close together in time should be more strongly correlated than observations taken further apart, and we would use a decreasing function of the time interval $\dt=|t_i-t_j|$ to represent that. One of the most widely used covariance functions is the squared exponential % nitpicky note: I just *love* Neil Lawrence's "Exponentiated square" much more than "squared exponential"
\begin{equation}
\label{eq:kse}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,\lengthscale^2}\right).
\end{equation}
This gives rise to smooth (infinitely differentiable) random functions with variance $\amplitude^2$ and characteristic length-scale $\lambda$. We present other commonly used kernel functions and discuss how to select one in \autoref{sec:kernels}.

@@

To address this, \citet{2009ApJ...704...51C} proposed a wavelet-based method to model correlated noise in transit light curves, which can be seen as a special case of GPR with a covariance function belonging to the exponential family, but formulated in terms of wavelets. Rather than modelling the covariance in the time domain, the Power Spectral Density (PSD) of the noise is assumed to be of the form $P(f) \propto f^{-\gamma}$.
Noting that the wavelet transform of such a process gives rise to a nearly diagonal covariance matrix, \citet{2009ApJ...704...51C}  derived an expression for the likelihood that can be computed in $\mathcal{O}(N)$ operations. 
This method has been used widely since its publication to model transit observations from space missions which stare continuously at a given field, such as \textit{Kepler} or TESS.
\begin{armarginnote}
  \entry{Colours of noise}{White, pink and red noise correspond to power spectral densities with power-law indices $\gamma=0$, $1$ and $2$, respectively.}
\end{armarginnote}
However, the use of a wavelet transform requires regular time sampling, which precludes its application to datasets with irregular sampling or significant data gaps, for example from ground-based, or space telescopes in low-Earth orbit such as Hubble.

@@

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/RV_framework.pdf}
  \caption{Schematic of the multidimensional GP framework of \citet{2015MNRAS.452.2269R} for modelling activity signals in RV data. The observed RVs (shown as orange points with error bars in the bottom right panel) are modelled alongside observations of activity indicators extracted from the spectra or line profiles (orange points with error bars in the bottom left and bottom middle panels). In all these time-series, the activity signal (shown by the black line in each panel) is modelled as a linear combination of an unobserved (latent) GP, shown in the top-left panel, and its first time-derivative, shown in the top middle panel. In addition, the Keplerian signal(s) of putative planet(s) (the sum of which is shown by the blue lines in the top right and bottom right panels) contribute to the RVs but not to the activity indicators. This hierarchical model is implemented in a Bayesian framework, sampling over all the hyperparameters (HP) listed in the blue boxes (subject to appropriate priors) to derive posteriors over the hyperparameters of interest and/or estimate the evidence for models with different numbers of planets. Models with different numbers of planets (or without planets) are then compared using the methods discussed in Section~\ref{sec:model_selection}.}
  \label{fig:RV_framework}
\end{figure}

@@

On the other hand, RV extraction pipelines routinely provide additional indicators measuring the characteristic width and asymmetry of the spectral lines, which are also affected by active regions, as well as spectral activity indicators which trace chromospheric emission in the cores of strong lines. \citet{2015MNRAS.452.2269R} developed a GP framework, schematically illustrated in \autoref{fig:RV_framework}, to model these activity indicators alongside the RVs, and to disentangle them from planetary signals. Each observed time-series is modelled as a linear combination of some latent variable $G(t)$, its time derivative $G'(t)$, white noise, and (in the case of the RVs only) one or more Keplerian signals. $G(t)$ does not have a direct physical interpretation; it loosely corresponds to $F^2$ in the $FF'$ framework of \citet{2012MNRAS.419.3147A}, so that $G'(t) \propto F(t) F'(t)$, but it is modelled as a (typically quasi-periodic) GP, which makes it possible to write down expressions for the covariance between any pair of observations from any of the time-series included in the analysis. This results in a global covariance matrix of dimensions $(NM,NM)$, where $N$ is the number of observations and $M$ the number of activity indicators included in the analysis (including the RVs themselves). This framework has proved particularly successful for RV confirmation of transiting planets around young stars \citep{2019MNRAS.490..698B,2022MNRAS.512.3060Z}, enabling the detection of planetary signals almost 50 times smaller than the activity signals.  Open-source implementations of this framework have been published \citep{2022MNRAS.509..866B,2022A&A...659A.182D}, and extensions thereof are likely to play a significant role in the analysis of next-generation RV surveys targeting Earth analogues, which need to achieve even better contrast between activity and planetary signals. \citet{2022arXiv220506627C} proposed a GP regression network model for stellar activity signals in RV, showing that it can be used to model RVs of the Sun-as-a-star simultaneously with other activity indicators.

@@

Given the current state of the field, our current recommendation for assessing and mitigating model misspecification is to use the methods discussed in \autoref{sec:model_selection} to determine the suitability of a GP for a specific application.
In the case where there are significant outliers in the dataset, the most mature and commonly used approach is to perform iteratively sigma clipping the data as a pre-processing step.
As an alternative, \citet{2015ApJ...812..128C} propose a method for handling non-stationary and structured model misspecification in the context of stellar spectral modelling, by designing a basis of ``local'' kernel functions tuned to capture line shape residuals.
Such domain-specific methods can be developed and validated using the methods discussed in \autoref{sec:model_selection}.

@@

It is reasonably straightforward to implement a simple GP model in code, and within astrophysics it has been common for authors to implement custom GPs for their analysis.
However, things get significantly more complicated when implementing the scalable or approximate methods discussed in the previous section.
Similarly, it can be tedious to experiment with different kernel functions and inference methods without building a non-trivial modelling infrastructure, something that has typically been done ad-hoc in astrophysics research.
